Pleiotropy of expression quantitative trait loci increases with the number of regulated genes, active cell types and tissues and bound transcription factors

### 2023-02-26 ms/ms03_results.tex
ENH ucsc tracks
MS

The colocalization pipeline uncovered eQTL colocalizations for 5,484 of the leading GWAS variants with 7,040 eQTL genes and 127 eQTL biosamples in the somatic chromosomes \cite{2019.Posthuma.Watanabe,2021.Li.Mu}.
%
% select count(*) from (select distinct tophits.chrom,tophits.pos,tophits.ea,tophits.rsid
% from tophits inner join coloc on coloc.rsid=tophits.rsid where coloc.pp_h4_abf>=0.75 and coloc.snp_pp_h4>=0.5) as sel
% 5484
% select count(*) from (select distinct tophits.chrom,tophits.pos,tophits.ea,tophits.rsid
% from tophits inner join coloc on coloc.rsid=tophits.rsid where coloc.pp_h4_abf>=0.75 and coloc.snp_pp_h4>=0.5) as sel
% 2441
% select count(*) from (select distinct eqtl_gene_id from coloc where pp_h4_abf>=0.75) as sel;  7040
% select count(*) from (select distinct eqtl_gene_id from coloc where pp_h4_abf>=0.75 and coloc.snp_pp_h4>=0.5) as sel;  5620
% select count(*) from (select distinct eqtl_id from coloc where pp_h4_abf>=0.75) as sel;  127
%
Variants in the MHC locus in chromosome 6 between 25,000,000 and 35,000,000 were
excluded from the colocalization pipeline, because of their complex linkage disequilibrium structure.
%
The histogram shows three groups of GWAS depending on the percentage of explained leading SNPs (Fig. \ref{fig:hist_perc_tophits_eqtl_excl_mhc}a).
%
For around 25\% of GWAS, there are not colocalizations between the leading SNPs and eQTLs.
%
On the other side, for around 7.5\% of GWAS, 100\% of the leading SNPs colocalize with eQTLs.
%
The remaining GWAS show between 10\% and 50\% colocalized leading SNPs with eQTLs.

The number of colocalization depends on the PP.H4.SNP parameter that gives the probability of a SNP in the colocalized locus to be the causal SNP.
%
We have changed this parameter between 25\% and 75\%, which results in 5\% and 12\% of the GWAS
showing colocalization with 100\% of their leading variants.
%
In this range, between 16\% and 25\% of the GWAS do not show any colocalization.
%
The remaining GWAS show between 5\% and 10\% of their leading variants colocalized with eQTLs (Fig. \ref{fig:hist_perc_tophits_eqtl_excl_mhc}a and Supplementary figure S1).

@@

I provide a UCSC track hub with these GWAS colocalizations.
%
In Fig. \ref{fig:hist_perc_tophits_eqtl_excl_mhc}b-d, I show three screenshots of this UCSC track hub for three different biosamples, namely B-cells, naive monocytes and frontal cortex biosamples in the chromosome 5 around the gene IL4.
%
This screenshot shows that depending on the biosample, the regulated genes are very different.
%
In the

@@

In the 417 GWAS, there are traits from different categories and belonging to independent studies.
%
To verify that the colocalization results are coherent between different GWAS datasets,
I cluster the GWAS traits based on the Spearman correlation between beta values
of colocalized eQTL variant, gene and biosample (Fig. \ref{fig:trait_clusters}).

@@

Then I wanted to verify if different traits cluster coherently in different trait classes.
%
As trait distance, I used the the spearman correlation of the eQTL splitted by
chromosome, position, alternative allele, eQTL gene and eQTL biosample.
%
To avoid non resolve eQTLs, I set a cutoff SNP.PP.H4>=0.5 in addition to the cutoff PP.H4.abf>=0.75.
%
The cutoff SNP.PP.H4>=0.5 was chosen, it allow to group pleiotropic eQTLs in a reasonable number of regions (See below).
%
I plotted these trait distances as a hierarchically-clustered heatmap.
%
Only trait classes with more than a minimum number of traits were plotted (Fig. \ref{fig:trait_clusters}).
%
Due to the large number of datasets, I have create a separate class for some cancers such as breast and skin.
%
In the OpenGWAS, some traits are redundant because the come from the same reference.
%
I plotted non-rendundant GWAS traits.
%
This analysis shows a coherent clustering of traits belonging to autoimmune, circulatory system diseases, cancers and allergies (Fig. \ref{fig:trait_clusters}).
%
Breast and skin cancer are found in two different cluster with some shared signal.
%
Multiple sclerosis cluster with autoimmune diseases of the digestive system.

@@

Next I verified whether the colocalized eQTL/GWAS variants tend to be specific to traits, eQTL genes and eQTL tissues.
%
74\% of colocalized eQTL/GWAS variants are associated to one GWAS class, 21\% are associated to 2 classes and the remaining variants associated to
3 or more trait classes make less that 4\% of colocalized eQTL/GWAS variants (Fig. \ref{fig:hist_gwas_egene_etissue}a).
%
Regarding eQTL genes, 36\% of the colocalized eQTL/GWAS variants modulate a single gene, 24\% modulate two genes, and the
remaining 40\% modulate three or more genes (Fig. \ref{fig:hist_gwas_egene_etissue}b).
%
Concerning tissues, 27\% colocalized eQTL/GWAS variants are active in a single tissue, 16\% in two tissues, 11\% in three
tissues, and the remaining 46\% in four or more tissues (Fig. \ref{fig:hist_gwas_egene_etissue}c).
%
In conclusion, trait pleiotropy is rather rare among colocalized eQTL/GWAS variants, while, regulation of at least two
genes in at least two tissues is rather common (Fig. \ref{fig:hist_gwas_egene_etissue}).

>>>>

The colocalization pipeline uncovered eQTL colocalizations for 5,484 of the leading GWAS variants with 7,040 eQTL genes and 127 eQTL biosamples in the somatic chromosomes \cite{2019.Posthuma.Watanabe,2021.Li.Mu}.
%
% select count(*) from (select distinct tophits.chrom,tophits.pos,tophits.ea,tophits.rsid
% from tophits inner join coloc on coloc.rsid=tophits.rsid where coloc.pp_h4_abf>=0.75 and coloc.snp_pp_h4>=0.5) as sel
% 5484
% select count(*) from (select distinct tophits.chrom,tophits.pos,tophits.ea,tophits.rsid
% from tophits inner join coloc on coloc.rsid=tophits.rsid where coloc.pp_h4_abf>=0.75 and coloc.snp_pp_h4>=0.5) as sel
% 2441
% select count(*) from (select distinct eqtl_gene_id from coloc where pp_h4_abf>=0.75) as sel;  7040
% select count(*) from (select distinct eqtl_gene_id from coloc where pp_h4_abf>=0.75 and coloc.snp_pp_h4>=0.5) as sel;  5620
% select count(*) from (select distinct eqtl_id from coloc where pp_h4_abf>=0.75) as sel;  127
%
Variants in the MHC locus in chromosome 6 between 25,000,000 and 35,000,000 were
excluded from the colocalization pipeline, because of their complex linkage disequilibrium structure.
%
The histogram shows three groups of GWAS depending on the percentage of explained leading SNPs (Fig. \ref{fig:1}a).
%
For around 25\% of GWAS, there are not colocalizations between the leading SNPs and eQTLs.
%
On the other side, for around 7.5\% of GWAS, 100\% of the leading SNPs colocalize with eQTLs.
%
The remaining GWAS show between 10\% and 50\% colocalized leading SNPs with eQTLs.

The number of colocalization depends on the PP.H4.SNP parameter that gives the probability of a SNP in the colocalized locus to be the causal SNP.
%
We have changed this parameter between 25\% and 75\%, which results in 5\% and 12\% of the GWAS
showing colocalization with 100\% of their leading variants.
%
In this range, between 16\% and 25\% of the GWAS do not show any colocalization.
%
The remaining GWAS show between 5\% and 10\% of their leading variants colocalized with eQTLs (Fig. \ref{fig:1}a and Supplementary figure S1).

@@

I provide a UCSC track hub with these GWAS colocalizations.
%
In Fig. \ref{fig:1}b-d, I show three screenshots of this UCSC track hub for three different biosamples, namely monocytes, and  frontal cortex biosamples in the chromosome 5 around the gene IL4 (Fig. \ref{fig:1}a).
%
This screenshot shows that depending on the biosample, the regulated genes are very different.
%
In the

@@

In the 417 GWAS, there are traits from different categories and belonging to independent studies.
%
To verify that the colocalization results are coherent between different GWAS datasets,
I cluster the GWAS traits based on the Spearman correlation between beta values
of colocalized eQTL variant, gene and biosample (Fig. \ref{fig:2}).

@@

Then I wanted to verify if different traits cluster coherently in different trait classes.
%
As trait distance, I used the the spearman correlation of the eQTL splitted by
chromosome, position, alternative allele, eQTL gene and eQTL biosample.
%
To avoid non resolve eQTLs, I set a cutoff SNP.PP.H4>=0.5 in addition to the cutoff PP.H4.abf>=0.75.
%
The cutoff SNP.PP.H4>=0.5 was chosen, it allow to group pleiotropic eQTLs in a reasonable number of regions (See below).
%
I plotted these trait distances as a hierarchically-clustered heatmap.
%
Only trait classes with more than a minimum number of traits were plotted (Fig. \ref{fig:2}).
%
Due to the large number of datasets, I have create a separate class for some cancers such as breast and skin.
%
In the OpenGWAS, some traits are redundant because the come from the same reference.
%
I plotted non-rendundant GWAS traits.
%
This analysis shows a coherent clustering of traits belonging to autoimmune, circulatory system diseases, cancers and allergies (Fig. \ref{fig:2}).
%
Breast and skin cancer are found in two different cluster with some shared signal.
%
Multiple sclerosis cluster with autoimmune diseases of the digestive system.

@@

Next I verified whether the colocalized eQTL/GWAS variants tend to be specific to traits, eQTL genes and eQTL tissues.
%
74\% of colocalized eQTL/GWAS variants are associated to one GWAS class, 21\% are associated to 2 classes and the remaining variants associated to
3 or more trait classes make less that 4\% of colocalized eQTL/GWAS variants (Fig. \ref{fig:3}a).
%
Regarding eQTL genes, 36\% of the colocalized eQTL/GWAS variants modulate a single gene, 24\% modulate two genes, and the
remaining 40\% modulate three or more genes (Fig. \ref{fig:3}b).
%
Concerning tissues, 27\% colocalized eQTL/GWAS variants are active in a single tissue, 16\% in two tissues, 11\% in three
tissues, and the remaining 46\% in four or more tissues (Fig. \ref{fig:3}c).
%
In conclusion, trait pleiotropy is rather rare among colocalized eQTL/GWAS variants, while, regulation of at least two
genes in at least two tissues is rather common (Fig. \ref{fig:3}).

### 2023-02-26 ms/ms06_backmatter.tex
MS

    \begin{figure}[h!]
        %
        \caption{Histogram showing the percentage of GWAS with the percentage of their tophits colocalized with eQTLs.
        I have excluded tophits in in the MHC locus in chromosome 6 between 25,000,000 and 35,000,000,
            because this locus is not included in the colocalization pipeline.}
        \label{fig:hist_perc_tophits_eqtl_excl_mhc}
        %
    \end{figure}

@@

    \begin{figure}[h!]
        %
        \caption{Trait clustering based on the Spearman correlation between beta values of colocalized eQTL variants, genes and biosamples.}
        \label{fig:trait_clusters}
        %
    \end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Fig 3: Histograms of variants vs GWAS, eQTL genes and eQTL biosamples
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
    \begin{figure}[h!]
        %
        \caption{Percentage of colocalized eQTL/GWAS variants with different number of GWAS classes (\textbf{a}), eQTL genes
            (\textbf{b}) and eQTL tissues (\textbf{c}).} \label{fig:hist_gwas_egene_etissue}
        %
    \end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Fig 4: Spearman correlation between counts of GWAS categories, eQTL genes and eQTL biosamples
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    \begin{figure}[h!]
        %
        \caption{Spearman correlation between the counts of GWAS classes, eQTL genes and eQTL biosamples.}
        \label{fig:correlation_counts}
        %
    \end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Fig 5: VEP consequences
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

    \begin{figure}[h!]
        %
        \caption{Odds ratio of the variants annotated with given variant effect
        predictor (VEP) consequence and given GWAS class count compared with non
        pleiotropic variants with GWAS class count one. The significance is given for the
        q-value with False discovery rate 5\%.}
        \label{fig:vep_consequence}
        %
    \end{figure}

@@

    \begin{figure}[h!]
        %
        \caption{TODO beta and p-val}
        \label{fig:beta_pval}
        %
    \end{figure}

@@

    \begin{figure}[h!]
        %
        \caption{Mean number of eQTL genes per eQTL-biosample pair (\textbf{a}) and mean number of biosamples per eQTL-gene pairs (\textbf{b}).} \label{fig:gwas_egene_etisue_per_variant}
        %
    \end{figure}

@@

    \begin{figure}[h!]
        %
        \caption{Analysis of transcription factor binding.
        \textbf{a}, Binding count of transcription factors in the region surrounding pleiotropic eQTLs with a radius of 50 bp as a function of the GWAS class count.
        \textbf{b}, Odds ratio of eQTLs annotated as belonging to a cis regulatory modules vs. non-annotated.}
        \label{fig:freq_tf_per_variant}
        %
    \end{figure}

>>>>

    \begin{figure}[!ht]
        %
        \caption{Histogram showing the percentage of GWAS with the percentage of their tophits colocalized with eQTLs.
        I have excluded tophits in in the MHC locus in chromosome 6 between 25,000,000 and 35,000,000,
            because this locus is not included in the colocalization pipeline.}
        \label{fig:1}
        %
    \end{figure}

@@

    \begin{figure}[!ht]
        %
        \caption{Trait clustering based on the Spearman correlation between beta values of colocalized eQTL variants, genes and biosamples.}
        \label{fig:2}
        %
    \end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Fig 3: Histograms of variants vs GWAS, eQTL genes and eQTL biosamples
% Fig 4: Spearman correlation between counts of GWAS categories, eQTL genes and eQTL biosamples
% Fig 5: VEP consequences
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
    \begin{figure}[!ht]
        %
        \caption{Distribution of GWAS category counts, correlation and ENSEMBL variant effect prediction.
        \textbf{a-c} Distribution of GWAS category, eQTL gene and eQTL biosample counts.
        Percentage of colocalized eQTL/GWAS variants with different number of GWAS classes (\textbf{a}), eQTL genes (\textbf{b}) and eQTL tissues (\textbf{c}).
        \textbf{d} Spearman correlation between the counts of GWAS categories, eQTL genes and eQTL biosamples.
        \textbf{e} Odds ratio of the eQTLs annotated with the given ENSEMBL variant predictor
        for eQTLs association withe the given GWAS category count versus the eQTLs with a single GWAS category count.
        The significance is given for the q-value with False discovery rate 5\%.
        }
        \label{fig:3}
        %
    \end{figure}

@@

    \begin{figure}[!ht]
        %
        \caption{eQTL and GWAS beta and p-values and allele frequencies for different GWAS category counts.
        \textbf{a,b} Average eQTL and GWAS beta values.
        \textbf{c,d} Average eQTL and GWAS p-values.
        \textbf{e} Average alternative allele frequency in the European population from the 1000 Genomes database.
        }
        \label{fig:4}
        %
    \end{figure}

@@

    \begin{figure}[!ht]
        %
        \caption{Mean number of eQTL genes per eQTL-biosample pair (\textbf{a}) and mean number of biosamples per eQTL-gene pairs (\textbf{b}).} \label{fig:gwas_egene_etisue_per_variant}
        %
    \end{figure}

@@

    \begin{figure}[!ht]
        %
        \caption{Analysis of transcription factor binding.
        \textbf{a}, Binding count of transcription factors in the region surrounding pleiotropic eQTLs with a radius of 50 bp as a function of the GWAS class count.
        \textbf{b}, Odds ratio of eQTLs annotated as belonging to a cis regulatory modules vs. non-annotated.}
        \label{fig:freq_tf_per_variant}
        %
    \end{figure}
